WASP-4 is Accelerating Towards the Earth

### 2020-01-24 paper/ms.tex
update paper

% TODO: may want to improve title
\title{WASP-4 is accelerating towards the Earth: A Cautionary Tale in
the Search for Tidal Orbital Decay}
% "An Explanation for WASP-4's Decreasing Orbital Period".
% "WASP-4's Decreasing Orbital Period: Solved."

@@

\begin{abstract}
  We recently noted that combining measurements from the Transiting
  Exoplanet Survey Satellite (TESS) with ground-based transit
  observations, the orbital period of WASP-4b seemed to be decreasing.
  Possible explanations included tidal orbital decay, orbital
  precession, and light-travel time effects.
  In this work we present new observations of the radial velocity of
  WASP-4, acquired with Keck-HIRES.
  They show that the system is accelerating towards our line of sight,
  at $\dot{\gamma} = X.XX \pm Y.YY\ {\rm m}\,{\rm s}^{-1}\,{\rm
  yr}^{-1}$.
  This implies a period decrease of X.XX milliseconds per year,
  compared to the observed value of Y.YY.
  Combining the radial velocities with new speckle imaging from
  Gemini-Zorro, we show that the acceleration is best explained by a
  10-200$\,M_{\rm Jup}$ companion between 10 and 100$\,$AU.
  We encourage further long-term monitoring of hot Jupiter radial
  velocities; these observations are critical in the ongoing hunt
  for tidal orbital decay.
\end{abstract}

@@

The present study highlights a point that, while obvious, has perhaps
not yet received due attention.  The point is that the hunt for
orbital decay of hot Jupiters will be crippled without long-term
radial velocity monitoring programs.
The reason is simple: when the timescale of orbital period change far
exceeds the observing baseline, the first deviation from constant
periodicity is always quadratic in the transit number:
\begin{align}
  t_{\rm tra} = t_0 + P E + \frac{1}{2}\frac{dP}{dE}E^2,
\end{align}
where
\begin{align}
  \frac{dP}{dE} = P \frac{dP}{dt},
\end{align}
$E$ is the transit number, $P$ is the orbital period, $t_0$ is the
reference epoch, and $t_{\rm tra}$ is the transt mid-time.

There are ample reasons to expect orbital decay to be common
(CITE).
However there are reasons to expect the R\'omer delay to perhaps be
even {\it more} common (CITE Knutson 14, Bryan 16).
Both manifest to first order identically in transit times (up to the
sign of the line-of-sight acceleration) -- as a non-zero period
derivative.

Specifically, \citet{knutson_friends_2014} showed that X\% of hot
Jupiters show some form of radial acceleration in radial
velocity time-series with baselines of YY years.
Assuming a separation distribution of (WHATEVER), this implies that
ZZ\% of hot Jupiters will {\it always} show a changing orbital period,
over timescale of say 10 years.

This is specific to hot Jupiters. The abundance of giant planets
exterior to Super-Earth sized systems is also high (CITE Bryan 2019).

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that from transits appears to be decreasing by about 10
milliseconds per year.
After discovering the timing variations in the TESS data 
\citep[][hereafter \citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}, we obtained N additional
radial velocity measurements using
Keck-HIRES, extending the RV baseline by Y years.
Previously, the radial velocities were sparse and were consistent with
not showing any linear trend.
Our new measurements reveal a line of sight acceleration of
$\dot{\gamma} = -XX {\rm m\,s^{-1}\,yr^{-1}}$.
This translates to an expected period decrease simply from the light
travel time effect of ZZ milliseconds per year --- about what is
observed.

Thereafter, \citet{southworth_transit_2019} reported 22 new transit
times for the system, and confirmed that the updated series of transit
times was consistent with a quadratic ephemeris.  Their interpretation
of the timing variations did not differ in any major respects from our
own, though with additional data they found a lower best-fit decay
rate of $\dot{P} = -XX.X \pm X.X$ milliseconds per year.

A separate study by \citet{baluev_homogeneously_2019} reported
additional light curves. \citeauthor{baluev_homogeneously_2019}
analyzed their newly obtained photometry, along with archival light
curves that we omitted from our analysis due to systematic
uncertainties in the absolute time system.
\citeauthor{baluev_homogeneously_2019} found that when they used all
the available TTV data, the need for a quadratic ephemeris was present
``at the high $\sim 5-7$ sigma level''.  However, they pointed out
that if they used lower-precision subsets of the available timing
data, the necessity for the quadratic term decreased.
\citeauthor{baluev_homogeneously_2019} also pointed out that the
precise transit times reported by \citet{huitson_gemini_2017} were
quite important in the time-series.  Overall,
\citeauthor{baluev_homogeneously_2019} did not find the claim of a
period decrease convincing.

% FIXME: figure out correct sections
\S~\ref{sec:observations} presents the new observations...
\S~\ref{sec:analysis} describes our analysis...
\S~\ref{sec:discussion} discusses...
\S~\ref{sec:conclusion} gives conclusions...

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.9\textwidth]{f1.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Timing residuals and best-fit models for WASP-4b.}
  The vertical axis shows the observed times minus the calculated
  times assuming a constant orbital period .  Darker points correspond
  to more precise data.  The quadratic ephemeris and its $1\sigma$
  uncertainties are shown in blue.  The binned TESS point (red square)
  is the weighted average of 18 TESS transits and is for display
  purposes only.  The models were fitted to all of the individual
  transit times.
  \label{fig:times}
	}
\end{figure*}

@@

%FIXME: is the purple here from my Pdot, or Southworth's?
%FIXME: perhaps include the WASP-4b phase-fold here too??
\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.48\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocity observations of WASP-4.} The best-fit Keplerian orbit
    of WASP-4b has been subtracted.  The linear trend inferred from the
    RV data is shown with a black line, with $1\sigma$ errors in gray.
    The trend that would be needed to produce the period decrease seen
    in transits ($\dot{P}= - YY.Y \pm Z.Z~{\rm ms}\,{\rm yr}^{-1}$) is
    indicated with the purple dotted line.  The four new RV
    measurements presented in this work increased the significance of
    the slope from $\approx$2$\sigma$ to 15$\sigma$.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure}

@@

%TODO: describe 3b better
\begin{figure*}[!t]
	\gridline{
		\fig{f3a.png}{0.33\textwidth}{}
		\fig{f3b.pdf} {0.5\textwidth}{}
	}
	\vspace{-0.7cm}
    \caption{
      {\it Left:\,} 
      Speckle image reconstructed from 1000 40 millisecond frames in an 832 nm bandpass, and acquired on September 28, 2019.
      The image scale is $2.46''\times2.46''$.
      {\it Right:\,} 
      Contrast curve derived from point-source injection-recovery
      experiments.
    }
    \label{fig:zorro}
\end{figure*}

@@

\subsection{How many other hot Jupiters will show the R\'omer delay?}

@@

To evaluate the importance of these effects in future transit timing
searches, we collected the linear trends reported by
\citet{knutson_friends_2014}, and computed the expected orbital period
changes $\dot{P}_{\rm RV} = \dot{v}_{\rm r} P/c$ for all the
systems.
The results are given in {\bf Table~X}, and visualized for hot
Jupiters with significant linear trends in Figure~\ref{fig:pdot_pop}.

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.48\textwidth]{f4.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf ...}
	\label{fig:pdot_pop}
  \vspace{-0.3cm}
	}
\end{figure}

@@

@@

\acknowledgements
%
This paper includes data collected by the TESS mission, which are
publicly available from the Mikulski Archive for Space Telescopes
(MAST).
%
Funding for the TESS mission is provided by NASA's Science Mission
directorate.
%
This work made use of NASA's Astrophysics Data System Bibliographic
Services.
%
Based on observations obtained at the Gemini Observatory, which is
operated by the Association of Universities for Research in Astronomy,
Inc., under a cooperative agreement with the NSF on behalf of the
Gemini partnership: the National Science Foundation (United States),
National Research Council (Canada), CONICYT (Chile), Ministerio de
Ciencia, Tecnolog\'{i}a e Innovaci\'{o}n Productiva (Argentina),
Minist\'{e}rio da Ci\^{e}ncia, Tecnologia e Inova\c{c}\~{a}o (Brazil),
and Korea Astronomy and Space Science Institute (Republic of Korea).
%
Observations in the paper made use of the High-Resolution Imaging
instrument Zorro at Gemini-South. Zorro was funded by the NASA
Exoplanet Exploration Program and built at the NASA Ames Research
Center by Steve B. Howell, Nic Scott, Elliott P. Horch, and Emmett
Quigley.
%
This research has made use of the VizieR catalogue access tool, CDS,
Strasbourg, France. The original description of the VizieR service was
published in A\&AS 143, 23.
%
This work has made use of data from the European Space Agency (ESA)
mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed
by the {\it Gaia} Data Processing and Analysis Consortium (DPAC,
\url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding
for the DPAC has been provided by national institutions, in particular
the institutions participating in the {\it Gaia} Multilateral
Agreement.
%
\newline
%
\software{
  \texttt{astrobase} \citep{bhatti_astrobase_2018},
	\texttt{astroplan} \citep{astroplan2018},
  \texttt{astropy} \citep{the_astropy_collaboration_astropy_2018},
  \texttt{astroquery} \citep{astroquery_2018},
  % \texttt{BATMAN} \citep{kreidberg_batman_2015},
  \texttt{corner} \citep{corner_2016},
  \texttt{emcee} \citep{foreman-mackey_emcee_2013},
  \texttt{IPython} \citep{perez_2007},
  \texttt{matplotlib} \citep{hunter_matplotlib_2007}, 
	\texttt{MESA} \citep{paxton_modules_2011,paxton_modules_2013,paxton_modules_2015}
  \texttt{numpy} \citep{walt_numpy_2011}, 
  \texttt{pandas} \citep{mckinney-proc-scipy-2010},
  \texttt{radvel} \citep{fulton_radvel_2018},
  % \texttt{scikit-learn} \citep{scikit-learn},
  \texttt{scipy} \citep{jones_scipy_2001}.
}

>>>>

% TODO: may want to improve title
\title{WASP-4 is Accelerating Towards the Earth}
% ": A Cautionary Tale in the Search for Tidal Orbital Decay"
% "An Explanation for WASP-4's Decreasing Orbital Period".
% "WASP-4's Decreasing Orbital Period: Solved."

@@

\begin{abstract}
  We recently reported that space and ground-based transit observations
  showed the orbital period of WASP-4b to be decreasing.
  Possible explanations included tidal orbital decay, orbital
  precession, and light-travel time effects.
  In this work we present new observations of the radial velocity of
  WASP-4, acquired with Keck-HIRES.
  The new data show that the system is accelerating towards our line
  of sight, at $\dot{\gamma} = X.XX \pm Y.YY\ {\rm m}\,{\rm
  s}^{-1}\,{\rm yr}^{-1}$.
  This implies a period decrease of X.XX milliseconds per year,
  compared to the observed value of Y.YY.
  Combining the radial velocities with new speckle imaging from
  Gemini-Zorro, we find that the WASP-4 system likely has a
  10-200$\,M_{\rm Jup}$ companion with semi-major axis between 10 and
  100$\,$AU.
  As observing baselines continue to extend, long-term tracking of hot
  Jupiter radial velocities will be necessary to distinguish tidal
  orbital decay from the R{\o}mer delay.
\end{abstract}

@@

The present study highlights a point that, while obvious, has perhaps
not yet received due attention.  The point is that the hunt for
orbital decay of hot Jupiters will be crippled without long-term
radial velocity monitoring programs.  The reason is simple: when the
timescale of orbital period change far exceeds the observing baseline,
the first deviation from constant periodicity is always quadratic in
the transit number.

There are ample reasons to expect orbital decay to be common (CITE).
However there are reasons to expect the R{\o}mer delay to perhaps be
even {\it more} common (CITE Knutson 14, Bryan 16).  Both manifest to
first order identically in transit times (up to the sign of the
line-of-sight acceleration) -- as a non-zero period derivative.

Specifically, \citet{knutson_friends_2014} showed that X\% of hot
Jupiters show some form of radial acceleration in radial velocity
time-series with baselines of YY years.  Assuming a separation
distribution of (WHATEVER), this implies that ZZ\% of hot Jupiters
will {\it always} show a changing orbital period, over timescale of
say 10 years.
The abundance of giant planets
exterior to Super-Earth sized systems is also high (CITE Bryan 2019),
but this is not so much of a concern to us, as their planetary masses
and therefore tidal timescales are typically expected to be at least
an order of magnitude shorter than hot Jupiters.

The main focus of this study is the hot Jupiter WASP-4b, which has an
orbital period that from transits appears to be decreasing by about 10
milliseconds per year.  We discovered the timing variations using
data from TESS and a decade of ground-based monitoring
\citep[][hereafter \citetalias{bouma_wasp4b_2019}]{bouma_wasp4b_2019}.

Thereafter, \citet{southworth_transit_2019} reported 22 new transit
times for the system, and confirmed that the updated series of transit
times was consistent with a quadratic ephemeris.  With additional
data, they found a lower best-fit decay rate of $\dot{P} = -XX.X \pm
X.X$ milliseconds per year.

A separate study by \citet{baluev_2019} reported
additional archival transit light curves of WASP-4b, most notably from
the TRAPPIST archive, and a select few other observers.
% \citeauthor{baluev_homogeneously_2019} analyzed their newly obtained
% photometry, along with archival light curves that we omitted from our
% analysis due to systematic uncertainties in the absolute time system.
\citeauthor{baluev_2019} found that when they used all
the available TTV data, the need for a quadratic ephemeris was present
``at the high $\sim 5-7$ sigma level''.  However, they pointed out
that if they used lower-precision subsets of the available timing
data, the necessity for the quadratic term decreased. This is not
particularly surprising.
% \citeauthor{baluev_homogeneously_2019} also pointed out that the
% precise transit times reported by \citet{huitson_gemini_2017} were
% quite important in the time-series.  Overall,
% \citeauthor{baluev_homogeneously_2019} did not find the claim of a
% period decrease convincing.

To determine the origin of the period change, we acquired four
additional radial velocity measurements using Keck-HIRES, extending
the RV baseline from 3 to 9 years.  Previously, the radial velocities
were sparse and showed a weak ($<2\sigma$) linear trend.  Our new
measurements reveal a line of sight acceleration of $\dot{\gamma} =
-XX {\rm m\,s^{-1}\,yr^{-1}}$.  This translates to an expected period
decrease simply from the light travel time effect of ZZ milliseconds
per year --- about commensurate with what is observed.

% FIXME: figure out correct sections
\S~\ref{sec:observations} presents the new observations...
\S~\ref{sec:analysis} describes our analysis...
\S~\ref{sec:discussion} discusses...
\S~\ref{sec:conclusions} gives conclusions...

@@

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f1.pdf}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Timing residuals and best-fit models for WASP-4b.}
  The vertical axis shows the observed times minus the calculated
  times assuming a constant orbital period .  Darker points correspond
  to more precise data.  The quadratic ephemeris and its $1\sigma$
  uncertainties are shown in blue.  The binned TESS point (red square)
  is the weighted average of 18 TESS transits and is for display
  purposes only.  The models were fitted to all of the individual
  transit times.
  \label{fig:times}
	}
\end{figure}

@@

%FIXME: is the purple here from my Pdot, or Southworth's?
%FIXME: perhaps include the WASP-4b phase-fold here too??
\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.5\textwidth]{f2.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
    {\bf Radial velocity observations of WASP-4.} The best-fit Keplerian orbit
    of WASP-4b has been subtracted.  The linear trend inferred from the
    RV data is shown with a black line, with $1\sigma$ errors in gray.
    The trend that would be needed to produce the period decrease seen
    in transits ($\dot{P}= - YY.Y \pm Z.Z~{\rm ms}\,{\rm yr}^{-1}$) is
    indicated with the purple dotted line.  The four new RV
    measurements presented in this work increased the significance of
    the slope from $\approx$2$\sigma$ to 15$\sigma$.
	\label{fig:rvs}
  \vspace{-0.3cm}
	}
\end{figure}

@@

%TODO: describe 3b better
%TODO: also, put 3b in as the bottom right inset to the contrast curve
\begin{figure}[!t]
	\gridline{
		\fig{f3a.png}{0.33\textwidth}{}
	}
	\vspace{-1cm}
	\gridline{
		\fig{f3b.pdf} {0.5\textwidth}{}
	}
	\vspace{-0.7cm}
    \caption{
      {\it Top:\,} 
      Speckle image reconstructed from 1000 40 millisecond frames in an 832 nm bandpass, and acquired on September 28, 2019.
      The image scale is $2.46''\times2.46''$.
      {\it Bottom:\,} 
      Contrast curve derived from point-source injection-recovery
      experiments.
    }
    \label{fig:zorro}
\end{figure}

@@

\subsection{How many other hot Jupiters will show the R{\o}mer delay?}

@@

To evaluate the importance of these effects in future transit timing
searches, we collected the linear trends reported by
\citet{knutson_friends_2014}, and computed the expected orbital period
derivatives $\dot{P}_{\rm RV} = \dot{v}_{\rm r} P/c$ for all the
systems.
The results are given in {\bf Table~X}, and visualized for hot
Jupiters with significant linear trends in Figure~\ref{fig:pdot_pop}.

\begin{figure}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.48\textwidth]{f5.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
  {\bf Predicted hot Jupiter period changes from linear radial
  velocity trends.} Including WASP-4b, 16 of 51 hot Jupiters from
  \citet{knutson_friends_2014} have shown long-term radial velocity
  trends.  The HAT-P-11 signal is plotted, though its radial velocity
  signal may be caused by stellar activity.  Three hot Jupiters are
  not plotted because their radial velocity curves better described as
  quadratic trends in time: HAT-P-17, WASP-8, and WASP-34.  Objects
  are ordered in the $y$ dimension by the absolute value of d$P$/d$t$.
	\label{fig:pdot_pop}
  \vspace{-0.3cm}
	}
\end{figure}

@@

We found that WASP-4 is accelerating towards the Earth, at
X.XX m/s/yr.
This seems like the most plausible explanation for the orbital 
period decrease observed in transits~---~not tidal
orbital decay, or apsidal precession.
The most likely parameters for the companion are XXX.
Over half of hot Jupiters have super-Jovian outer companions,
so the R{\o}mer delay will become an increasingly large nuisance in
the hunt for orbital decay as
the observing baselines for hot Jupiters increase.
The requirement to avoid missing out is to have long-term
radial velocity monitoring with the same spectrograph (or else to
calibrate instrumental offsets by observing the same stars simultaneously).

%TODO: what's the scaling? sigma_P ~ 1/baseline? baseline^2?

@@

% \acknowledgements
% %
% This paper includes data collected by the TESS mission, which are
% publicly available from the Mikulski Archive for Space Telescopes
% (MAST).
% %
% Funding for the TESS mission is provided by NASA's Science Mission
% directorate.
% %
% This work made use of NASA's Astrophysics Data System Bibliographic
% Services.
% %
% Based on observations obtained at the Gemini Observatory, which is
% operated by the Association of Universities for Research in Astronomy,
% Inc., under a cooperative agreement with the NSF on behalf of the
% Gemini partnership: the National Science Foundation (United States),
% National Research Council (Canada), CONICYT (Chile), Ministerio de
% Ciencia, Tecnolog\'{i}a e Innovaci\'{o}n Productiva (Argentina),
% Minist\'{e}rio da Ci\^{e}ncia, Tecnologia e Inova\c{c}\~{a}o (Brazil),
% and Korea Astronomy and Space Science Institute (Republic of Korea).
% %
% Observations in the paper made use of the High-Resolution Imaging
% instrument Zorro at Gemini-South. Zorro was funded by the NASA
% Exoplanet Exploration Program and built at the NASA Ames Research
% Center by Steve B. Howell, Nic Scott, Elliott P. Horch, and Emmett
% Quigley.
% %
% This research has made use of the VizieR catalogue access tool, CDS,
% Strasbourg, France. The original description of the VizieR service was
% published in A\&AS 143, 23.
% %
% This work has made use of data from the European Space Agency (ESA)
% mission {\it Gaia} (\url{https://www.cosmos.esa.int/gaia}), processed
% by the {\it Gaia} Data Processing and Analysis Consortium (DPAC,
% \url{https://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding
% for the DPAC has been provided by national institutions, in particular
% the institutions participating in the {\it Gaia} Multilateral
% Agreement.
%
% \newline
%

\software{
  \texttt{astrobase} \citep{bhatti_astrobase_2018},
	\texttt{astroplan} \citep{astroplan2018},
  \texttt{astropy} \citep{the_astropy_collaboration_astropy_2018},
  \texttt{astroquery} \citep{astroquery_2018},
  % \texttt{BATMAN} \citep{kreidberg_batman_2015},
  \texttt{corner} \citep{corner_2016},
  \texttt{emcee} \citep{foreman-mackey_emcee_2013},
  \texttt{IPython} \citep{perez_2007},
  \texttt{matplotlib} \citep{hunter_matplotlib_2007}, 
	\texttt{MESA} \citep{paxton_modules_2011,paxton_modules_2013,paxton_modules_2015}
  \texttt{numpy} \citep{walt_numpy_2011}, 
  \texttt{pandas} \citep{mckinney-proc-scipy-2010},
  \texttt{radvel} \citep{fulton_radvel_2018},
  % \texttt{scikit-learn} \citep{scikit-learn},
  \texttt{scipy} \citep{jones_scipy_2001}.
}
